Local Spectral Clustering of Density Upper Level Sets

### 2019-05-23 neurips/local_spectral_density_clustering_neurips.tex
fixed a few typos; final version

\begin{abstract}
We analyze Personalized PageRank (PPR), a local spectral method for clustering,
which extract clusters using locally-biased random walks around a user-specified
seed node.  In constrast to pervious work, we adopt a traditional statistical
learning setup, where we obtain samples from an unknown distribution, and aim to
identify connected regions of high-density (density clusters).  We prove that
PPR, run on a neighborhood graph, extracts sufficiently salient density
clusters. We also provide empirical support for our theory.
\end{abstract}

@@

To motivate these conditions, \citet{zhu2013} show for arbitrary graph $G =
(V,E)$ and subset of vertices $S \subseteq V$, the PPR estimate \smash{$\Cest$}
of subset $S$ satisfies, for a constant $c>0$,
\begin{equation}
\label{eqn: graph_symmetric_set_difference_1}
\vol(\Cest \setminus S; G) + \vol(S \setminus \Cest; G) \leq c \bigl( \Phi(S,G)
\cdot \tau_{\infty}(G[S])\bigr) \vol(S;G),
\end{equation}
where $\Phi(S;G)$ is the normalized cut of $S$ (as defined in \eqref{eqn:
  normalized_cut}), and \smash{$\tau_{\infty}(G[S])$} is called the \emph{mixing
  time} of a random walk over the induced subgraph $G[S]$ (to be defined
precisely later, in \eqref{eqn: mixing_time}).  The left-hand side in
\eqref{eqn:  graph_symmetric_set_difference_1} resembles a (degree-weighted) 
form of the symmetric set difference metric in \eqref{eqn:
  misclassification_rate}.  As we will show in Section \ref{sec:
  analysis}, the conditions \ref{asmp: bounded_density}--\ref{asmp:
  bounded_volume} allow us to upper bound the normalized cut 
\smash{$\Phi(\Csig[\Xbf]; G_{n,r})$}, and the mixing time 
\smash{$\tau_{\infty}(G_{n,r}[\Csig[\Xbf]])$}.\footnote{Informally, assumptions  
\ref{asmp: cluster_separation}, \ref{asmp: low_noise_density} yield an upper
bound on \smash{$\cut(\Csig[\Xbf]; G_{n,r})$}, and \ref{asmp: bounded_density}
yields a lower bound on \smash{$\vol(\Csig[\Xbf]; G_{n,r})$}; together with 
\ref{asmp: bounded_volume}, this gives an upper bound on the normalized cut.  On 
the other hand, \ref{asmp: bounded_density}, \ref{asmp: embedding} preclude
bottlenecks in the induced subgraph \smash{$G_{n,r}[\Csig[\Xbf]]$}, and combined
with the upper bound in \ref{asmp: embedding}, this leads to an upper bound on
the mixing time over this subgraph.}

@@

The proof of Theorem \ref{thm: mixing_time_upper_bound} relies heavily on
analogous mixing time bounds developed for the mixing time of the
continuous-space ball walk over convex sets. To the best of our knowledge, our
result is the first bound, albeit asymptotic, on the mixing time of random walks
over neighborhood graphs that is independent of $n$, the number of vertices.

@@

\paragraph{Empirical behavior of PPR.} In Figure \ref{fig:fig2}, to drive home
the main implications of Theorems \ref{thm: misclassification_rate} and
\ref{thm: consistent_recovery_of_density_clusters}, we show the
behavior of PPR, normalized cut, and the density clustering algorithm of
\citet{chaudhuri2010} on the well known ``two moons'' dataset (with added 2d 
Gaussian noise), considered a prototypical success story for spectral clustering
algorithms. The first column shows the empirical density clusters $C[X]$
and $C'[X]'$ for a particular threshold $\lambda$ of the density function; the 
second column shows the cluster recovered by PPR; the third column shows the
global minimum normalized cut, computed according to the algorithm of
\citet{szlam2010}; and the last column shows a cut of the density cluster tree
estimator of \citet{chaudhuri2010}.  We can see the degrading ability of PPR to
recover density clusters as the two moons become less well-separated. Of 
particular interest is the fact that PPR fails to recover one of the moons even
when normalized cut still succeeds in doing so, supporting our claim from Remark
\ref{rmk: diameter}. Additionally, we note that the Chaudhuri-Dasgupta algorithm 
succeeds even when both PPR and normalized cut fail.  While our main message was
that PPR recovers geometrically well-conditioned density clusters, it would be  
interesting to establish that it \emph{only} recovers such clusters, a direction
for future work.

>>>>

\begin{abstract}
We analyze Personalized PageRank (PPR), a local spectral method for clustering,
which extract clusters using locally-biased random walks around a user-specified
seed node.  In constrast to previous work, we adopt a traditional statistical
learning setup, where we obtain samples from an unknown distribution, and aim to
identify connected regions of high-density (density clusters).  We prove that
PPR, run on a neighborhood graph, extracts sufficiently salient density
clusters. We also provide empirical support for our theory.
\end{abstract}

@@

To motivate these conditions, \citet{zhu2013} show for arbitrary graph $G =
(V,E)$ and subset of vertices $S \subseteq V$, the PPR estimate \smash{$\Cest$}
of subset $S$ satisfies, for a constant $c>0$,
\begin{equation}
\label{eqn: graph_symmetric_set_difference_1}
\vol(\Cest \setminus S; G) + \vol(S \setminus \Cest; G) \leq c \bigl( \Phi(S,G)
\cdot \tau_{\infty}(G[S])\bigr) \vol(S;G),
\end{equation}
where $\Phi(S;G)$ is the normalized cut of $S$ (as defined in \eqref{eqn:
  normalized_cut}), and \smash{$\tau_{\infty}(G[S])$} is called the \emph{mixing
  time} of a random walk over the induced subgraph $G[S]$ (to be defined
precisely later, in \eqref{eqn: mixing_time}).  The left-hand side in
\eqref{eqn:  graph_symmetric_set_difference_1} resembles a (degree-weighted) 
form of the symmetric set difference metric in \eqref{eqn:
  misclassification_rate}.  As we will show in Section \ref{sec:
  analysis}, the conditions \ref{asmp: bounded_density}--\ref{asmp:
  bounded_volume} allow us to upper bound the normalized cut 
\smash{$\Phi(\Csig[\Xbf]; G_{n,r})$}, and the mixing time 
\smash{$\tau_{\infty}(G_{n,r}[\Csig[\Xbf]])$}.\footnote{Informally, assumptions  
\ref{asmp: cluster_separation}, \ref{asmp: low_noise_density} yield an upper
bound on \smash{$\cut(\Csig[\Xbf]; G_{n,r})$}, and \ref{asmp: bounded_density}
yields a lower bound on \smash{$\vol(\Csig[\Xbf]; G_{n,r})$}; together with 
\ref{asmp: bounded_volume}, this gives an upper bound on the normalized cut.  On 
the other hand, \ref{asmp: bounded_density}, \ref{asmp: embedding} preclude
bottlenecks in the induced subgraph \smash{$G_{n,r}[\Csig[\Xbf]]$}, and combined
with the upper bound on diameter in \ref{asmp: embedding}, this leads to an upper bound on
the mixing time over this subgraph.}

@@

The proof of Theorem \ref{thm: mixing_time_upper_bound} relies heavily on
analogous mixing time bounds developed for the mixing time of a
continuous-space ``ball walk'' over convex sets. To the best of our knowledge, our
result is the first bound, albeit asymptotic, on the mixing time of random walks
over neighborhood graphs that is independent of $n$, the number of vertices.

@@

\paragraph{Empirical behavior of PPR.} In Figure \ref{fig:fig2}, to drive home
the main implications of Theorems \ref{thm: misclassification_rate} and
\ref{thm: consistent_recovery_of_density_clusters}, we show the
behavior of PPR, normalized cut, and the density clustering algorithm of
\citet{chaudhuri2010} on the well known ``two moons'' dataset (with added 2d 
Gaussian noise), considered a prototypical success story for spectral clustering
algorithms. The first column shows the empirical density clusters $\Cset[\Xbf]$
and $\Cset'[\Xbf]$ for a particular threshold $\lambda$ of the density function; the 
second column shows the cluster recovered by PPR; the third column shows the
global minimum normalized cut, computed according to the algorithm of
\citet{szlam2010}; and the last column shows a cut of the density cluster tree
estimator of \citet{chaudhuri2010}.  We can see the degrading ability of PPR to
recover density clusters as the two moons become less well-separated. Of 
particular interest is the fact that PPR fails to recover one of the moons even
when normalized cut still succeeds in doing so, supporting our claim from Remark
\ref{rmk: diameter}. Additionally, we note that the Chaudhuri-Dasgupta algorithm 
succeeds even when both PPR and normalized cut fail.  While our main message was
that PPR recovers geometrically well-conditioned density clusters, it would be  
interesting to establish that it \emph{only} recovers such clusters, a direction
for future work.
